Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International \emph{and Domestic} Students

### 2024-02-22 demystify.tex
updates

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system (i.e., not by the CS department).  If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't, then the system will inform you what is missing and what you need to do.

@@

Then, applications are reviewed by a \textbf{PhD admission committee} (\emph{adcom}) that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that GMU, UMass, and probably many other schools, have separate adcom for MS programs.

Each application is assigned to about \emph{three} faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are likely the main ones deciding your application, \emph{every} faculty in the department will have access to your application and can provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA, see \autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths (e.g., a well-known school) and weaknesses (e.g., weak
LORs).

@@

>>>>

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system, i.e., not by the CS department. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that many schools, including GMU, have separate adcoms for MS programs.

Each application is assigned to 2--3 faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are likely the main ones deciding your application, other faculty in the department will likely have access to your application and can provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. Some members can be tenured faculty. The chair of the committee will likely a senior faculty, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA, see \autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.

@@

\section{Why did I get rejected?}
Many possible reasons!  Other than the obvious ones (e.g., weak research pontentials, low GPA, weak LoRs, unclear SOP), here are some that you might not expect:
\begin{itemize}
  \item  \emph{Your English scores (e.g., IELTS or TOEFL) are not high enough}.  While professors do not care much about these, there is a certain minimum set by the college or the university that you need to pass to get admitted, especially if you want to get funding (e.g., you need to be able to communicate well with students as a TA).  Note this is mostly for international students, as domestic one do not have to take these tests.

  \item \emph{No faculty available to advise you}.  You could have an excellent profile, but if there is no faculty who can advise you (e.g., they already have too many students), you will not be admitted. This is actually for your own good because you do not want to be in a program where you cannot find an adviser.
\end{itemize}

\section{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We cannot reveal your evaluation results (e.g., if we say because you have weak LORs, then that would reveal something about your writers) or give you advice on how to improve. So it is best to just move on.  If you really want feedback, ask your professors, collaborators, ref writers, or those who have previously applied.
